---
title: "Lab 06 — Hypothesis Testing from Claims"
author: "EAES 480 — Modern Statistics in Earth & Environmental Science"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: cosmo
    highlight: tango
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    df_print: paged
editor_options:
  chunk_output_type: inline
---

> GM GRADE 10/10 - Very strong submission. Clean code, excellent visualizations (including creative use of violin + axis break), thoughtful scientific interpretations, and good EAES connections. Minor deductions for slightly awkward hypothesis formatting and a very low temperature threshold for the "cold" group.

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(janitor)

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)
```

# Overview

This lab introduces **hypothesis testing** as an extension of sampling, bootstrap distributions, and uncertainty quantification.

The core idea is simple:

> A **claim** about the population becomes a **null hypothesis**.

You will use the **US-AMS simplified AmeriFlux dataset** again and test claims about population means using:

-   one sample,
-   a bootstrap estimate of uncertainty,
-   a z-statistic,
-   and a p-value.

This lab emphasizes **reasoning from claims to tests**, not just mechanics.

------------------------------------------------------------------------

# Learning goals

By the end of this lab, you should be able to:

-   Turn a verbal claim into a null hypothesis
-   Choose a variable and define the relevant population parameter
-   Draw a sample and estimate a mean
-   Use bootstrap resampling to estimate uncertainty in that mean
-   Compute a z-statistic and p-value
-   Interpret a hypothesis test in plain language
-   Connect confidence intervals and hypothesis tests

------------------------------------------------------------------------

# Data

## Load and inspect

```{r load_data, echo=TRUE, eval=TRUE}
# GOAL: Read the dataset and clean the names.

df <- read_csv("data/us-ams-simple.csv", na = c("-9999")) %>%
  clean_names() %>%
  filter(year_local == 2023)

glimpse(df)
```

## Derive time columns (if needed)

```{r derive_time, echo=TRUE, eval=TRUE}
# GOAL: Create date / month / day_night columns as in earlier labs.

df <- df %>%
  mutate(
    date = as.Date(floor(doy) - 1, origin = paste0(year_local, "-01-01")),
    month = month(date, label = TRUE, abbr = TRUE),
    day_night = if_else(daytime == 1, "Day", "Night")
  )
```

------------------------------------------------------------------------

# Part I — From claim to hypothesis test

## 1) Choose a variable

Choose **one numeric variable** from the dataset.

Possible examples: - `swc` - `ta` - `fc` - `gpp` - `le`

```{r choose_variable, echo=TRUE, eval=TRUE}
target_var <- "le"

# CHECK: does this variable exist?
target_var %in% names(df)
```

## 2) Describe the population parameter

Write the population parameter you care about:

> $\mu$ = the true 2023 population mean of `latent heat flux`

Fill in the blank in words:

> In this lab, my population parameter is the true annual mean of **`latent heat flux`** at the US-AMS site.

------------------------------------------------------------------------

## 3) Make a claim

Write a claim about the population mean for your variable.

Examples: - “The annual mean soil water content is 0.30.” - “The annual mean air temperature is 10 °C.” - “The annual mean latent heat flux is 40 W m$^{-2}$.”

Now turn your claim into hypotheses:

-   Null hypothesis: $H_0: \mu = \mu_0$
-   Alternative hypothesis: $H_A: \mu \neq \mu_0$

Fill these in:

> My claim is: `The annual mean latent heat flux is 100 W m$^{-2}$.`

> $H_0: \mu = \mu_0$ - The annual mean latent heat flux is not significantly different from 100 W m$^{-2}$.

> $H_A: \mu \neq \mu_0$ - The annual mean latent heat flux is significantly different from 100 W m$^{-2}$.

------------------------------------------------------------------------

## 4) Draw one sample and compute the sample mean

```{r draw_sample, echo=TRUE, eval=TRUE}
# GOAL: Draw one random sample and compute its mean.

set.seed(100)

n_samp <- 200   # e.g., 200

samp <- df %>%
  filter(!is.na(.data[[target_var]])) %>%
  slice_sample(n = n_samp)

sample_mean <- mean(samp[[target_var]], na.rm = TRUE)

sample_mean

```

**Prompt (2–3 sentences):** How does your sample mean compare to your claim? Is it close, or already quite different?

> *It's kind of close, considering the range of values we're working with. Latent heat flux has a range of ~550 units, so being only ~40 units off isn't too bad. It's not terribly different.*

------------------------------------------------------------------------

## 5) Bootstrap the sample mean

```{r bootstrap_means, echo=TRUE, eval=TRUE}
# GOAL: Bootstrap the sample mean to estimate uncertainty.

set.seed(100)

B <- 4000   # e.g., 4000

boot_means <- replicate(
  B,
  mean(sample(samp[[target_var]], size = n_samp, replace = TRUE), na.rm = TRUE)
)

head(boot_means)
```

## 6) Estimate uncertainty

```{r bootstrap_summary, echo=TRUE, eval=TRUE}
# GOAL: Compute the bootstrap SE and 95% CI.

boot_se <- sd(boot_means)
boot_ci <- quantile(boot_means, probs = c(0.025, 0.975))

tibble(
  sample_mean = sample_mean,
  boot_se = boot_se,
  ci_lower = boot_ci[1],
  ci_upper = boot_ci[2]
)
```

**Prompt (2–3 sentences):** What does the bootstrap SE tell you? What does the 95% CI tell you?

> *The bootstrap SE tells us the standard deviation of the 4000 bootstrap samples. The 95% CI tells us that 95% of the time, it is expected the 'true' value falls between 52.384 and 76.611. *

------------------------------------------------------------------------

## 7) Convert the claim into a formal test

Define the null value from your claim.

```{r hypothesis_test, echo=TRUE, eval=TRUE}
# GOAL: Compute z and p-value for your claim.

mu0 <- 100   # numeric claim value

z_stat <- (sample_mean - mu0) / boot_se

p_value <- 2 * pnorm(-abs(z_stat))

tibble(
  sample_mean = sample_mean,
  mu0 = mu0,
  boot_se = boot_se,
  z_stat = z_stat,
  p_value = p_value
)
```

------------------------------------------------------------------------

## 8) Interpret the result

Use your p-value and CI to answer:

-   Does the 95% CI include the null value?
-   Is the p-value small or large?
-   Would you reject the null hypothesis at $\alpha = 0.05$?

Fill in:

> The null value `100` is not inside my 95% CI. :(

> My p-value is `7.309943e-09`, which is small.

> Therefore, I would reject $H_0$ at the 0.05 level.

**Prompt (4–6 sentences):** Explain your decision in plain language. Do the data make the claim look plausible or surprising? Be careful: “fail to reject” does not mean “prove true.”

> *I said that the average latent heat flux is 100 Watts / square meter. However, the population mean is ~63 units and the 95% confidence interval is 52.38 - 76.61. This gives me a z-score of -5.78, which tells me 2 things: 1) The magnitude is large, meaning that my guess (100) is quite `off`, and 2) my guess is an overshoot (not undershoot) since the z score is negative.*
> *These data make my claim look surprising. Therefore, I reject the null hypothesis - the data do not support it.*

> GM: Excellent — clear, plain-language reasoning and correct caution about failing to reject.

------------------------------------------------------------------------

## 9) Visualize the hypothesis test

```{r plot_test, echo=TRUE, eval=TRUE}
# GOAL: Plot the bootstrap distribution, sample mean, CI, and null value.

ggplot(tibble(boot_mean = boot_means), aes(x = boot_mean)) +
  geom_histogram(bins = 100, alpha = 0.8) +
  geom_vline(xintercept = sample_mean, linetype = "dashed", linewidth = 1.1) +
  geom_vline(xintercept = boot_ci, color = "red", linewidth = 1.1) +
  geom_vline(xintercept = mu0, color = "blue", linewidth = 1.1) +
  theme_classic(base_size = 18) +
  labs(
    x = paste0("Bootstrap mean of ", target_var),
    y = "Count",
    title = "Bootstrap distribution and null claim",
    subtitle = "Dashed = sample mean; Red = 95% CI; Blue = null value"
  )
```

**Prompt (2–3 sentences):** Does the plot agree with your p-value-based conclusion? Explain.

> *Yes. It shows my guessed mean is outside the 95% CI and above the true mean (which is why the z-score was negative).*

> GM: Excellent visual confirmation.

------------------------------------------------------------------------

## 10) Try a second claim (short extension)

Keep the same variable, but test a **different null value**.

```{r second_claim, echo=TRUE, eval=TRUE}
# GOAL: Test a second claim using the same sample and bootstrap SE.

mu0_alt <- 50

z_stat_alt <- (sample_mean - mu0_alt) / boot_se
p_value_alt <- 2 * pnorm(-abs(z_stat_alt))

tibble(
  mu0_alt = mu0_alt,
  z_stat_alt = z_stat_alt,
  p_value_alt = p_value_alt
)
```

**Prompt (3–4 sentences):** How did changing the null value change the conclusion? What does this tell you about the relationship between the null hypothesis and the observed estimate?

> *I made the null value smaller (and closer to the population mean). This changed the sign of the z-score, decreased the z-score magnitude, and increased the p-value. We can no longer reject the null hypothesis; 50 could be the true pop. mean.*
>*This tells me that the null hypothesis value (50) is closer to the observed estimate (63).*

> GM: Excellent demonstration of sensitivity to the null value.

------------------------------------------------------------------------

# Part II — Comparing group means with t-tests

In Part I, you tested a claim about **one population mean**.

Now you will extend the same logic to compare **two group means**.
-   comparing means between groups,
-   one- vs two-sided tests,
-   and connections to t-tests / ANOVA.

The key idea is:

> A two-sample t-test asks whether the difference between two group means is large relative to the uncertainty in that difference.

You will first work with a built-in grouping variable (**day vs night**), then choose a second grouping variable of your own.

---

## 11) Create a day/night grouping variable

If you already created `day_night` earlier, you can reuse it. Otherwise, create it here.

```{r make_daynight, echo=TRUE, eval=TRUE}
# GOAL: create a day/night grouping variable from `daytime`.

df <- df %>%
  mutate(
    day_night = if_else(daytime == 1, "Day", "Night")
  ) %>% 
  filter(!is.na(day_night))

df %>% count(day_night)
```

---

## 12) Choose a response variable for the t-test

Pick one numeric response variable to compare between groups.

Possible examples:
- `ta`
- `swc`
- `fc`
- `gpp`
- `le`

```{r choose_response_ttest, echo=TRUE, eval=TRUE}
response_var <- "gpp"

# CHECK: does this variable exist?
response_var %in% names(df)
```

---

## 13) Visualize the response by day/night

Create a plot that shows how your response variable differs between Day and Night.

You may use:
- `geom_boxplot()`
- `geom_violin()`
- `geom_jitter()`
- or a combination

```{r plot_daynight, echo=TRUE, eval=TRUE}
# GOAL: visualize the response by day/night group.

df %>%
  filter(!is.na(.data[[response_var]])) %>%
  ggplot(aes(x = day_night, y = gpp, fill = day_night)) +
  geom_violin(alpha = 0.8) +
#  geom_boxplot(width=0.1, color="grey", alpha=0.2) + 
  theme_classic(base_size = 18) +
  labs(
    x = NULL,
    y = response_var,
    title = "Response by day vs night"
  )
```

**Prompt (2–3 sentences):** Before running the test, what do you predict?  
Do the groups look different in center? In spread?

> *Write your answer here.*

---

## 14) Summarize the two groups

Compute the group means, SDs, and sample sizes.

```{r summarize_daynight, echo=TRUE, eval=TRUE}
# GOAL: summarize the response variable by day/night group.

group_summary <- df %>%
  filter(!is.na(.data[[response_var]])) %>%
  group_by(day_night) %>%
  summarise(
    mean_response = mean(gpp),
    sd_response = sd(gpp),
    n = n(),
    .groups = "drop"
  )

group_summary
```

---

## 15) State the hypotheses for the day/night comparison

Write hypotheses in words and symbols.

Use this structure:

- Null: the two group means are equal
- Alternative: the two group means are different

Fill in:

> My response variable is `gpp`

> $H_0$: `The average GPP does not significantly change between day and night in 2023 in Argonne Nat'l Laboratory`

> $H_A$: `The average GPP changes significantly between day and night in 2023 in Argonne Nat'l Laboratory`

---

## 16) Run a two-sample t-test

Use `t.test()` to compare the means of your response variable between Day and Night.

```{r ttest_daynight, echo=TRUE, eval=TRUE}
# GOAL: run a two-sample t-test by formula.

t.test(
  gpp ~ day_night,
  data = df
)
```

**Prompt (4–6 sentences):** Report the p-value and confidence interval.  
Would you reject the null hypothesis at $\alpha = 0.05$?  
What does the result imply about day vs night differences in your chosen variable?

> *The p value is <0.01. The 95% CI is 10.91032 to 11.42790. I would reject the null hypothesis and instead state that the mean GPP at night and during the day at Argonne in '23 are significantly different.*

> GM: Excellent — strong reporting of p-value, CI, decision, and scientific meaning.
---

## 17) Connect the t-test to a confidence interval

Extract the confidence interval and interpret it as a difference in means.

```{r ttest_store, echo=TRUE, eval=TRUE}
# GOAL: store the t-test result in an object.

tt_daynight <- t.test(
  gpp ~ day_night,
  data = df
)

tt_daynight$conf.int
```

**Prompt (2–3 sentences):** Does the confidence interval include 0?  
How does that connect to the hypothesis test decision?

> *The confidence interval does not include 0. A CI that contains 0 indicates that the result is not statistically signficant. We do not have this issue, which supports the hypothesis test results.*

> GM: Excellent, direct and accurate.

---

## 18) Bootstrap view of the same comparison (conceptual extension)

Create bootstrap distributions of the **group means** for Day and Night separately.

```{r bootstrap_group_means, echo=TRUE, eval=TRUE}
set.seed(100)

B <- 2000   # e.g., 2000

x_day <- df %>%
  filter(day_night == "Day") %>%
  pull(gpp) %>%
  na.omit()

x_night <- df %>%
  filter(day_night == "Night") %>%
  pull(gpp) %>%
  na.omit()

boot_day <- replicate(
  B,
  mean(sample(x_day, size = length(x_day), replace = TRUE))
)

boot_night <- replicate(
  B,
  mean(sample(x_night, size = length(x_night), replace = TRUE))
)

boot_groups <- bind_rows(
  tibble(group = "Day", mean_est = boot_day),
  tibble(group = "Night", mean_est = boot_night)
)
```

```{r plot_boot_groups, echo=TRUE, eval=TRUE}
# GOAL: visualize the bootstrap distributions by group.
library(ggbreak)

ggplot(boot_groups, aes(x = mean_est, fill = group)) +
  geom_histogram(alpha = 0.45, bins = 1000, position = "identity") +
  theme_classic(base_size = 18) +
  labs(
    x = paste0("Bootstrap mean of ", response_var),
    y = "Count",
    title = "Bootstrap distributions of group means", 
    subtitle = "Discontinuous x-axis"
  ) + 
  scale_x_break(c(0.5, 10.5)) + 
  theme(plot.subtitle = element_text(face = "italic"))
```

**Prompt (3–4 sentences):** How does this bootstrap visualization help you understand the t-test result?  
Do the plausible group means strongly overlap, or are they clearly separated?

> *I appreciate visualizations in addition to statistical results; I think they're often critical to correct interpretation. There is no overlap. They're so clearly separated that I added in an x axis break so you could see the means. Without the axis break you couldn't see anything.*

---

## 19) Choose a second grouping variable of your own

Now define **one additional grouping variable** and repeat the t-test workflow.

Possible ideas:
- month grouped into two seasons (e.g., winter vs summer)
- a high/low split of another continuous variable
- a wet/dry split of SWC
- a daytime subset grouped by month category

You must:
- define the grouping variable clearly
- explain why the grouping is scientifically meaningful

```{r choose_second_group, echo=TRUE, eval=TRUE}
# GOAL: create your own two-level grouping variable.

df <- df %>%
  mutate(
    temp = ifelse(ta < 1.66667, "cold", "warm")
  )

df %>% count(temp)
```

**Prompt (2–3 sentences):** Why did you choose this grouping variable? What scientific contrast are you testing?

> *I'm doing "cold" and "warm" - when ambient temperature is < 40º F (~4.44 ºC) it's 'cold', and above this threshold it's 'warm'. I'm curious if GPP is lower on cooler days and higher on warmer days. It would probably be best to test temperature as a continuous variable, but I'm curious to see results.*

> GM: The cold/warm threshold of ~1.67°C is very low and limits the contrast. A higher threshold or full seasonal grouping would be stronger.

---

## 20) Visualize and test your second grouping variable

Repeat the workflow:
1. visualize
2. summarize
3. state hypotheses
4. run `t.test()`
5. interpret

```{r visualize_second_group, echo=TRUE, eval=TRUE}
# GOAL: visualize your response by your chosen grouping variable.

df %>% ggplot(aes(x = temp, y = gpp, fill = temp)) + 
  geom_violin() + 
  theme_bw()
  
```

```{r summarize_second_group, echo=TRUE, eval=TRUE}
# GOAL: summarize your response variable by your chosen grouping variable.
group_summary <- df %>%
  filter(!is.na(.data[[response_var]])) %>%
  group_by(temp) %>%
  summarise(
    mean_response = mean(gpp),
    sd_response = sd(gpp),
    n = n(),
    .groups = "drop"
  )

group_summary
```

```{r ttest_second_group, echo=TRUE, eval=TRUE}
# GOAL: run a two-sample t-test for your chosen grouping variable.

t.test(
  gpp ~ temp,
  data = df
)
```

**Prompt (5–6 sentences):** Compare the day/night t-test to your second grouping-variable test.  
Which contrast was stronger? Which was more scientifically meaningful?  
Did the two tests produce similar or different conclusions?

> *Both the day/night and warm/cool tests have p values less than 2.2e-16. There's probably some correlation between these values that I didn't think of until now - it's usually colder at night and warmer during the day. The results I'm seeing from temperature could be caused by temporal differences or vice versa. The conclusions of the two tests are similar, likely because of this correlation. Neither of these results are super meaningful because the values are temporally autocorrelated - GPP is going to be low at night b/c there's no photosynthesis. *

> GM: Excellent thoughtful comparison, including recognition of autocorrelation between day/night and temperature.

---

## 21) Final reflection

Answer in **4–6 sentences**:

- How is a two-sample t-test similar to the one-mean hypothesis test from Part I?
- What changes when we move from “testing one mean” to “testing a difference in means”?
- How did the bootstrap group-mean plot help your understanding?

> *Both the two-sample t-test and one-mean hypothesis test seek to see how different two classes of values are. When we test one mean, we only have to consider one 'group' of data. When we test two groups of data, we have to consider differences in data structure. Otherwise, it's pretty similar - you're testing how different two values are. I'm a big fan of pairing visualizations with statistics and I believe there's (almost) always a benefit to having a visualization alongside a statistical output. The statistical output and the visualization are both saying the same thing but in two different ways, which enhances understanding. *

> GM: Very good synthesis and emphasis on the value of pairing visualizations with statistics. I completely agree! (:

------------------------------------------------------------------------

# Submission

-   Knit your `.Rmd` to HTML.
-   Commit and push both the `.Rmd` and knitted `.html` to your GitHub repo.
-   Submit the GitHub repository link on Canvas.
